Appendix 3: Estimating species interactions

### 2015-10-26 Harris_2015/manuscript-materials/draft.Rmd
moar stuff

\setlength{\parskip}{3pt}

@@

**Abstract:** Estimating species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that is
capable of outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can also be estimated from observed co-occurrence
rates via maximum likelihood. Using simulated landscapes with known pairwise
interaction strengths, I evaluated Markov networks and six existing approaches.
The Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions
largely overpowered them. Two computationally efficient approximations, based on
controlling for indirect effects with linear or generalized linear models,
also performed well. Indirect effects reliably caused a common null modeling
approach to produce incorrect inferences, however.

@@

To the extent that nontrophic species interactions (such as competition) affect
community assembly, ecologists might expect to find signatures of these interactions in
species composition data [@macarthur_population_1958; @diamond_island_1975].
Despite decades of work and several major controversies, however
[@lewin_santa_1983; @strong_ecological_1984;
@connor_checkered_2013], existing methods for detecting competition's effects
on community structure are unreliable [@gotelli_empirical_2009]. In particular,
species' effects on one another can become lost in the complex web of
indirect effects. For example, the competitive interaction between the two shrub
species in Figure 1A can become obscured by their shared tendency to occur in
unshaded areas (Figure 1B). While ecologists have long known that indirect
effects can overwhelm direct ones at the landscape level
[@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our
methods for drawing inferences from observational data do not control for these effects
[e.g. @diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009;
@veech_probabilistic_2013; @pollock_understanding_2014]. To the extent that
indirect interactions like those in Figure 1 are generally important
[@dodson_complementary_1970], existing methods will thus not generally provide
much evidence regarding species' direct effects on one another. The goal of this
paper is to resolve this long-standing problem.

@@

#### Markov networks.
Markov networks provide a framework for translating back and forth between the
conditional (all-else-equal) relationships among species (Figure 1C) and the
kinds of species assemblages that these relationships produce. Here, I show how
a set of conditional relationships can be used to determine how groups of species
can co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),$

@@

\noindent where the second sum is over all $\binom{n}{2}$ species pairs. In this model,  $\alpha_{i}$ is
an intercept term determining the amount that the presence
of species $i$ contributes to the log-probability of $\vec{y}$; it directly
controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
that the co-occurrence of species $i$ and species $j$ contributes to the
log-probability; it controls the conditional relationship between two species,
i.e. the probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). For example, $\beta_{ij}$
might have a value of $+2$ for two mutualists, indicating that the odds of
observing one species are $e^2$ times higher in sites where its partner is
present than in comparable sites where its partner is absent. Because the
relative probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when negatively-associated
species co-occur, the model tends---all else equal---to produce assemblages that
have many pairs of positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then species'
overall co-occurrence patterns can be more complicated. Determining the
marginal relationships between species from their conditional interactions
entails summing over the different possible assemblages (Figure 2B). This
becomes intractable when the number of possible assemblages is large, though
several methods beyond the scope of this paper can be employed to keep the
calculations feasible [@lee_learning_2012; @salakhutdinov_learning_2008].
Alternatively, as noted below, some common linear and generalized linear methods
can also be used as computationally efficient approximations to the full network
[@lee_learning_2012; @loh_structure_2013].

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average [i.e. maximizing the likelihood ensures that the sufficient
statistics of the model distribution match the sufficient statistics of the data;
@murphy_machine_2012]. I used the rosalia package [@harris_rosalia_2015] for the
R programming language [@r_core_team_r_2015] to optimize the Markov network
parameters. The package was named after Santa Rosalia, the patron saint of
biodiversity, whose mythos played an important rhetorical role in the null model
debates of the 1970's and 1980's [@lewin_santa_1983].

#### Simulated landscapes.
I simulated several sets of landscapes using known parameters as a test bed for
different statistical methods. The first set of landscapes included the three
competing species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling from a probability distribution defined by
the figure's interaction coefficients (Appendix 1). Each of the methods
described below was then evaluated on its ability to correctly infer that the
two shrub species competed with one another, despite their frequent co-occurrence.

I then generated landscapes with up to 20 interacting species at 25, 200, or
1600 sites using three increasingly complex models (50 replicates for each
combination of size and model). As described in Appendix [[X]], I randomly drew
the “true” coefficient values for each replicate landscape in these simulations
so that most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.

The first set of 20-species landscapes, like the landscapes with three species,
were generated directly from a Markov network to ensure that the model could
recover the parameters used to generate the "observed" co-occurrence data.
Then, I added two environmental factors that varied from location to location
across the simulated landscapes, and simulated a new set of co-occurrence data
so that species' $\alpha$ coefficients depended on these environmental variables.
The latter set of simulated landscapes provide an important test of the methods’
ability to distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for co-occurrence
patterns---the inference procedure did not have access to any information about
the environmental or spatial variables that helped shape the landscape [cf
@connor_checkered_2013].  Finally, I simulated the population dynamics of 20
interacting species, as described in Appendix [[X]]

#### Recovering species interactions from simulated data.

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
models,  as described above. For the analyses with 20 species, I added a weak
regularizer (equivalent to a logistic prior with scale 2) to ensure that the
model estimates were always finite [[Appendix]].

I also evaluated six alternative methods: five from the existing literature,
plus a novel combination of two of these methods. The first alternative interaction
metric was the sample correlation between species' presence-absence vectors,
which summarizes their marginal association.  Next, I used partial correlations,
which summarize species' conditional relationships. This approach is
common in molecular biology [@friedman_sparse_2008], but less so in ecology
[see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions].
In the context of non-Gaussian data, the partial correlation can be thought of
as a computationally efficient approximation to the full Markov network model
[@loh_structure_2013]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014].

@@

#### Evaluating model performance.
For the simulated landscapes based on Figure 1, I assessed whether each method's
test statistic indicated a positive or negative relationship between the two
shrubs (Appendix 1). For the null model (Pairs), I calculated statistical
significance using its $Z$-score. For the Markov network, I used the Hessian
matrix to generate approximate confidence intervals and noted whether these
intervals included zero.

I then evaluated the relationship between each method's estimates and the "true"
interaction strengths among all of the species pairs from the larger simulated
landscapes. This determined which of the methods provide a consistent way to
know how strong species interactions are---regardless of which species were
present in a particular data set or how many observations were taken. To ensure
that the different test statistics (e.g. correlations versus $Z$ scores) were on
a common scale, I rescaled them using linear regression through the origin. After
rescaling each method’s estimates, I calculated squared errors between the scaled
interaction estimates and “true” interaction values across the simulated data
sets. These squared errors determined the proportion of variance explained
for different combinations of model type and landscape size (compared with a
null model that assumed all interaction strengths to be zero).

@@

#### Three species.
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the correlation
between these species was positive in 94% of replicates, and the
randomization-based null model falsely reported positive associations 100% of
the time. Worse, more than 98% of these false conclusions were statistically
significant. The partial correlation and Markov network estimates, on the other
hand, each correctly isolated the direct negative interaction between the shrubs
from their positive indirect interaction 94% of the time (although the
confidence intervals overlapped zero in most replicates).

#### Twenty species.
Despite some variability across contexts (Figure 3), the four methods that
controlled for indirect effects (the Markov network, the generalized linear
models, and the two partial correlation-based methods) clearly outperformed those
that did not.  Of these, the Markov network consistently performed the best,
especially on the smaller landscapes [which are also the ones that are most
representative of typical analyses in this field; @gotelli_empirical_2009].

Of the methods that did not control for indirect effects, Figure 3 shows that
simple correlation coefficients provided a more reliable indicator of species'
true interaction strengths than either the joint species distribution model
(BayesComm) or the null model (Pairs). Interestingly, the estimates from these
three approaches were tightly correlated [[Figure 3B]]. This suggests that the
effect sizes from the null model (and, to a lesser extent, the correlation
matrices from joint species distribution models) only contain noisy versions of
the same information that could be obtained more easily and interpretably by
calculating correlation coefficients between species' presence-absence vectors.

Bootstrap resampling indicated that the above ranking of the different methods
was robust (Appendix 3). In particular, the 95% confidence interval of the
bootstrap distribution indicated that the Markov network's overall $R^2$ value
was between [[14 and 18]] percent higher than the second-most effective
method (generalized linear models) and between [[2.12 and 2.38]] times higher than
could be achieved by the null model (Pairs).

@@

Markov networks---particularly the Ising model for binary networks---have been
studied for nearly a century [@cipra_introduction_1987], and the models'
properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species interactions
would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
the states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012], which would provide a
rigorous and straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species distribution
models [@pollock_understanding_2014], leading to a
better understanding of the interplay between biotic and abiotic effects on
community structure. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow one species to affect the sign or
strength of the relationship between two other species, tipping the balance
between facilitation and exploitation [@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species interactions. Null
and neutral models can be useful for clarifying our thinking about the numerical
consequences of species' richness and abundance patterns
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
null model must be interpreted with care [@roughgarden_competition_1983]. Even
in small networks with three species, it may simply not be possible to implicate
individual species pairs or specific ecological processes like competition by
rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
test statistic is effectively just a correlation coefficient (Figure 3B).
Simultaneous estimation of multiple ecological parameters seems like a much more
promising approach: to the extent that the models' relative performance on real
data sets is similar to the range of results shown in Figure 3A, scientists in
this field could often double their explanatory power by switching from null
models to Markov networks (or increase it substantially with linear or
generalized linear approximations). Regardless of the methods ecologists
ultimately choose, controlling for indirect effects could clearly improve our
understanding of species' direct effects on one another and on community
structure.

#### Acknowledgements:
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, C. F. Dormann, and two anonymous reviewers provided
very helpful feedback on the text.

@@

**Figure 1: A.** A small network of three competing species. The tree (top) tends not to
co-occur with either of the two shrub species, as indicated by the strongly
negative coefficient linking them. The two shrub species also compete with one
another, but more weakly (circled coefficient). **B.** In spite of the
competitive interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors positively
correlated (circled). **C.** Controlling for the tree species' presence with a
conditional (all-else-equal) approach such as a partial covariance or a
Markov network leads to correct identification of the negative shrub-shrub
interaction (circled).

**Figure 2: A.** A small Markov network with two species, defined by its $\alpha$ and
$\beta$ values.  The abiotic environment favors the occurrence of each species
($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
$\beta$ coefficient linking these two species implies that they co-occur less
than expected under independence. **B.** These coefficients determine the
probabilities of all four possible presence-absence combinations for Species 1
and Species 2. $\alpha_1$ is added to the exponent whenever Species 1 is present
($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the exponent
includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
ensures that the four probabilities sum to 1.  In this case, $Z$ is
about 18.5.  **C.** The expected frequencies of all possible
co-occurrence patterns between the two species of interest, as calculated in the
previous panel. **D.** If, as assumed in null model simulations, $\beta_{12}$
equaled zero (e.g. if the species no longer competed for the same resources),
then the reduction in competition would allow each species to increase its
overall occurrence rate.

**Figure 3: A.** Proportion of variance in interaction coefficients explained by each
method versus number of sampled locations.
**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
species can be predicted using the correlation between the presence-absence
vectors of those same species and from the number of sites on the landscape.

@@

\setlength{\parskip}{0pt}
\setlength{\parindent}{-1em}
\setlength{\leftskip}{1em}

>>>>

\setlength{\parskip}{1pt}

@@

**Abstract:** Estimating species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that is
capable of outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can also be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known pairwise interaction strengths, I evaluated Markov
networks and six existing approaches. The Markov networks consistently
outperformed other methods, correctly isolating direct interactions between
species pairs even when indirect interactions or abiotic factors largely
overpowered them. Two computationally efficient approximations, based on
controlling for indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling approach
to produce incorrect inferences, however.

@@

To the extent that nontrophic species interactions (such as competition) affect
community assembly, ecologists might expect to find signatures of these
interactions in species composition data [@macarthur_population_1958;
@diamond_island_1975]. Despite decades of work and several major controversies,
however [@lewin_santa_1983; @strong_ecological_1984; @connor_checkered_2013],
existing methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
another can become lost in the complex web of indirect effects. For example, the
competitive interaction between the two shrub species in Figure 1A can become
obscured by their shared tendency to occur in unshaded areas (Figure 1B). While
ecologists have long known that indirect effects can overwhelm direct ones at
the landscape level [@dodson_complementary_1970; @levine_competitive_1976], the
vast majority of our methods for drawing inferences from observational data do
not control for these effects [e.g. @diamond_island_1975;
@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
@pollock_understanding_2014]. To the extent that indirect interactions like
those in Figure 1 are generally important [@dodson_complementary_1970], existing
methods will thus not provide much evidence regarding species interactions.

@@

**Markov networks.**
Markov networks provide a framework for translating back and forth between the
conditional (all-else-equal) relationships among species (Figure 1C) and the
kinds of species assemblages that these relationships produce. Here, I show how
a set of conditional relationships can be used to determine how groups of species
can co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

$\displaystyle{p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),}$

@@

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$ species. In this
model, $\alpha_{i}$ is an intercept term determining the amount that the presence
of species $i$ contributes to the log-probability of $\vec{y}$; it directly
controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
that the co-occurrence of species $i$ and species $j$ contributes to the
log-probability; it controls the conditional relationship between two species,
i.e. the probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). For example, $\beta_{ij}$
might have a value of $+2$ for two mutualists, indicating that the odds of
observing one species are $e^2$ times higher in sites where its partner is
present than in comparable sites where its partner is absent.

Because the relative probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when negatively-associated
species co-occur, the model tends---all else equal---to produce assemblages that
have many pairs of positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).
Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then species'
overall co-occurrence patterns can be more complicated, and may require summing
over the different possible assemblages (Figure 2B) to determine the overall
outcomes.

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average [i.e., maximizing the likelihood ensures that the sufficient
statistics of the model distribution match the sufficient statistics of the data;
@murphy_machine_2012]. I used the rosalia package [@harris_rosalia_2015] for the
R programming language [@r_core_team_r_2015] to optimize the Markov network
parameters.

\noindent \textbf{Simulated landscapes.}
I simulated several sets of landscapes using known parameters as a test bed for
different statistical methods. The first set of landscapes included the three
competing species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling from a probability distribution defined by
the figure's interaction coefficients (Appendix 1). Each of the methods
described below was then evaluated on its ability to correctly infer that the
two shrub species competed with one another, despite their frequent co-occurrence.

I then generated landscapes with up to 20 interacting species at 25, 200, or
1600 sites using three increasingly complex models (50 replicates for each
combination of size and model). As described in Appendix [[X]], I randomly drew
the “true” coefficient values for each replicate landscape
so that most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.

The first set of 20-species landscapes, like the landscapes with three species,
were generated directly from a Markov network to ensure that the model could
recover the parameters used to generate the "observed" co-occurrence data.
Then, I added two environmental factors that varied from location to location
across the simulated landscapes, and simulated a new set of co-occurrence data
so that species' $\alpha$ coefficients depended on the local environment.
The latter set of simulated landscapes provide an important test of the methods’
ability to distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for co-occurrence
patterns---the inference procedure did not have access to any information about
the environmental or spatial variables that helped shape the landscape [cf
@connor_checkered_2013].  I generated the final set of landscapes by simulating
the population dynamics of 20 interacting species, as described in Appendix [[X]].

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
models,  as described above. For the analyses with 20 species, I added a weak
regularizer (equivalent to a logistic prior with scale 2) to ensure that the
model estimates were always finite [[Appendix]].

I also evaluated six alternative methods: five from the existing literature,
plus a novel combination of two of these methods. The first alternative interaction
metric was the sample correlation between species' presence-absence vectors,
which summarizes their marginal association.  Next, I used partial correlations,
which summarize species' conditional relationships. This approach is
common in molecular biology [@friedman_sparse_2008], but is rare in ecology
(see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions).
In the context of non-Gaussian data, the partial correlation can be thought of
as a computationally efficient approximation to the full Markov network model
[@loh_structure_2013]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014].

@@

\noindent \textbf{Evaluating model performance.}
For the simulated landscapes based on Figure 1, I assessed whether each method's
test statistic indicated a positive or negative relationship between the two
shrubs (Appendix 1). For the null model (Pairs), I calculated statistical
significance using its $Z$-score. For the Markov network, I used the Hessian
matrix to generate approximate confidence intervals and noted whether these
intervals included zero.

I then evaluated the relationship between each method's estimates and the "true"
interaction strengths among all of the species pairs from the larger simulated
landscapes. This determined which of the methods provide a consistent way to
know how strong species interactions are---regardless of which species were
present in a particular data set or how many observations were taken. To ensure
that the different test statistics (e.g. correlations versus $Z$ scores) were on
a common scale, I rescaled them using linear regression through the origin. After
rescaling each method’s estimates, I calculated squared errors between the scaled
interaction estimates and "true" interaction values across the simulated data
sets. These squared errors determined the proportion of variance explained
for different combinations of model type and landscape size (compared with a
null model that assumed all interaction strengths to be zero).

@@

\noindent \textbf{Three species.}
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the correlation
between these species was positive in 94% of replicates, and the
randomization-based null model falsely reported positive associations 100% of
the time. Worse, more than 98% of these false conclusions were statistically
significant. The partial correlation and Markov network estimates, on the other
hand, each correctly isolated the direct negative interaction between the shrubs
from their positive indirect interaction 94% of the time (although the
confidence intervals overlapped zero in most replicates).

\noindent
\textbf{Twenty species.}
Despite some variability across contexts (Figure 3), the four methods that
controlled for indirect effects (the Markov network, the generalized linear
models, and the two partial correlation-based methods) outperformed those
that did not.  Of these, the Markov network consistently performed the best,
especially on the smaller landscapes [which are also the most
representative of typical analyses in this field; @gotelli_empirical_2009].
As shown in Appendix [[X]], the standard errors associated with the estimates
in Figure 3 are all small (less than 0.01), so differences across methods should
not be attributed to sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows that
simple correlation coefficients provided a more reliable indicator of species'
true interaction strengths than either the joint species distribution model
(BayesComm) or the null model (Pairs). The estimates from these
three approaches were tightly correlated (Figure 3B), suggesting that the
effect sizes from the null model (and, to a lesser extent, the correlation
matrices from joint species distribution models) only contain noisy versions of
the same information that could be obtained more easily and interpretably by
calculating correlation coefficients between species' presence-absence vectors.

Whereas Figure 3 summarizes the accuracy of each model's point estimates, Figure
4 compares the inferential value of the Markov network versus the null model.
When the "true" $\beta$ coefficient is near zero, the Markov network's 95%
confidence intervals usually include zero (but not quite 95% of the time), and
it is rare for the model to confidently attribute the wrong sign to a
coefficient (Figure 4A). The null model, by contrast, typically rejects the null
hypothesis at the $p=0.05$ level regardless of the underlying parameters, and
the sign of its $Z$-scores is not a reliable indicator of the interaction's
actual sign.

@@

Markov networks---particularly the Ising model for binary networks---have been
studied for nearly a century [@cipra_introduction_1987], and the models'
properties, capabilities, and limits are well-understood. Using the same
framework for species interactions would thus allow ecologists to tap into an
enormous set of existing discoveries and techniques for dealing with indirect
effects, stability, and alternative stable states. Numerous other extensions are
possible: for example, the states of the interaction network can be modeled as a
function of the local abiotic environment [@lee_learning_2012], which would
provide a rigorous and straightforward approach to the difficult and important
task of incorporating whole networks of biotic interactions into species
distribution models [@pollock_understanding_2014], leading to a
better understanding of the interplay between biotic and abiotic effects on
community structure. Alternatively, models could allow one species to affect the
interactions between two other species [@tjelmeland_markov_1998], tipping the
balance between facilitation and exploitation [@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species interactions. Null
and neutral models can be useful for clarifying our thinking about the numerical
consequences of species' richness and abundance patterns
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
null model must be interpreted with care [@roughgarden_competition_1983]. Even
in small networks with three species, it may simply not be possible to implicate
individual species pairs or specific ecological processes like competition by
rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
test statistic is effectively just a correlation coefficient [[Figure 3B]].
When a non-null backdrop is not controlled for, Type I error rates can
skyrocket, the apparent sign of the interaction can change, and null models
can lead to reliably incorrect inferences (Figure 1, Figure 4).

Controlling for indirect effects via simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the range
of results shown in Figure 3A, scientists in this field could more than double
their explanatory power by switching from null models to Markov networks (or
increase it nearly as much with linear or generalized linear approximations).
Regardless of the methods ecologists ultimately choose, controlling for indirect
effects could clearly improve our understanding of species' direct effects on
one another and on community structure.

\noindent \textbf{Acknowledgements:}
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, C. F. Dormann, and two anonymous reviewers provided
very helpful feedback on the text.

@@

\noindent \textbf{Figure 1: A.} A small network of three competing species. The tree (top) tends not to
co-occur with either of the two shrub species, as indicated by the strongly
negative coefficient linking them. The two shrub species also compete with one
another, but more weakly (circled coefficient). **B.** In spite of the
competitive interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors positively
correlated (circled). **C.** Controlling for the tree species' presence with a
conditional (all-else-equal) approach such as a partial covariance or a
Markov network leads to correct identification of the negative shrub-shrub
interaction (circled).

\noindent \textbf{Figure 2: A.} A small Markov network with two species, defined by its $\alpha$ and
$\beta$ values.  The abiotic environment favors the occurrence of each species
($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
$\beta$ coefficient linking these two species implies that they co-occur less
than expected under independence. **B.** These coefficients determine the
probabilities of all four possible presence-absence combinations for Species 1
and Species 2. $\alpha_1$ is added to the exponent whenever Species 1 is present
($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the exponent
includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
ensures that the four probabilities sum to 1.  In this case, $Z$ is
about 18.5.  **C.** The expected frequencies of all possible
co-occurrence patterns between the two species of interest, as calculated in the
previous panel. **D.** If, as assumed in null model simulations, $\beta_{12}$
equaled zero (e.g. if the species no longer competed for the same resources),
then the reduction in competition would allow each species to increase its
overall occurrence rate.

\noindent \textbf{Figure 3: A.} Proportion of variance in interaction
coefficients explained by each method versus number of sampled locations.
**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
species can be predicted using the correlation between the presence-absence
vectors of those same species and from the number of sites on the landscape.

\noindent \textbf{Figure 4: A.} Inferences from the Markov network's 95%
confidence intervals on the "no-environment" simulated landscapes, as a function
of the "true" coefficient values. **B.** Inferences from the $Z$-statistics
estimated by the "Pairs" software for null modeling.

@@

\setlength{\parindent}{-1em}
\setlength{\leftskip}{1em}
